Survey of Cache Oblivious B-trees

### 2011-12-11 paper.tex
Changing the column width

\caption{An illustration of the dynamic COSB-tree structure}
\label{fig:string_structure}
\end{figure}
\subsection{Motivation}
In order for B-trees to remain versatile, they must be able to accommodate various forms of data. However, traditional B-trees and string B-trees have suboptimal performance when variable length strings are used as keys. In addition, string B-trees do not support compression, and commonly used front compression is handled sub-optimally. Regular B-trees and string B-trees are not cache oblivious, and thus incur poorer performance as described previously. Range queries have poor performance in B-trees. In \cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache Oblivious String B-tree (COSB-tree), which they argue works efficiently in these problem areas for variable length string keys. They also provide a modified front-compression scheme, which gives their approach the advantages of using less space and having improved locality within the compressed data.

Two experiments are performed, in which a regular B-tree implementation tuned to various block sizes is compared to a static COSB-tree implementation. The experiment involved measuring the amount of time required to perform: 440,000 and 450,000 inserts of random values, a range query of all values, and 1000 random searches. The results of the experiment show that COSB-trees outperformed B-trees most of the time for random searches, range queries and random inserts. However, some point between 440,000 and 450,000 inserts the COSB-tree had to rebalance the entire tree, causing it to fall behind in this measurement. However, Bender et al. argue that the experiment was biased towards the B-tree implementation, since the trees are "young", and thus still well organized. The COSB-tree implementation is also compared to Berkeley DB **cite berkeley db**, and achieves performance comparable to Berkeley DB with tuned parameters. The results seem to show that there is a window in which large and small-block size B-trees perform poorly, and that COSB-trees may be a reasonable alternative to improve performance.
\subsection{Description}

\subsubsection{General COSB-tree Structure}
The static COSB-tree consists of 2 layers: a centroid tree at the top, and an array of the keys stored in lexicographic order on the bottom. The leaves of the centroid tree contain pointers to keys in the array. The centroid tree is structured such that a search through the centroid tree identifies the local area of the desired key in the array. Then a sequential scan is performed across the array to find the key.

@@

\caption{A centroid tree example}
\label{fig:centroid}
\end{figure}
In this approach to solving the dictionary matching problem *cite dictionary matching problem paper*, a dictionary $D$ of $N$ keys is preprocessed using hashing and divide-and-conquer via the construction of a centroid tree over the keys. This technique forms the basis for the Cache Oblivious String B-tree (COSB-tree) design. Consider the compacted trie $T$ of $D$. Bender et al. observe that there exists a centroid vertex $\rho$ in $T$ such that $\rho$ has between N/3 and 2N/3 descendants, inclusive. When searching for a key $k$ in the compacted trie, you traverse the tree as follows: for each node $y$ that you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on into the trie rooted at $y$ or \emph{down trie} of $y$. Otherwise traverse the trie resulting from excluding $y$ and its subtree. The \emph{centroid tree} is obtained by rooting the tree at \emph{centroid vertex} $\rho$, and making its children recursively defined centroid trees of its up and down tries.

@@

\subsubsection{Modifying Front Compression to Preserve Locality}
This section provides a brief summary of how front compression is modified to optimize compression and decoding for static and dynamic cache-oblivious string B-trees. Consider a sequence of $j$ keys to store $k_{1}$,  $k_{2}$, ...,  $k_{j}$. Without compression, this requires $\sum_{j}||k_{j}||$ space in memory. To compress these keys using front compression, let $\pi_{j}$ be the longest common prefix between $k_{j}$ and $k_{j+1}$. Now instead we store the keys in the following format:  
\begin{center}
$k_{1}$,$||\pi_{2}||$,$\sigma_{2}$,$||\pi_{3}||$$\sigma_{3}$,...,$||\pi_{j}||$,$\sigma_{j}$
\end{center}
where $\sigma_{j}$ is the suffix of $k_{j}$ after removing the first $\pi_{j}$ bits. This saves almost $\sum_{j}||\pi_{j}||$ memory. However, since decoding $k_{j}$ involves concatenating the first $\pi_{j}$ bits of $k_{j-1}$ to $\sigma_{j}$, $k_{j-1}$ must be decoded. $k_{j-1}$ may depend on previous keys as well, making $k_{j}$ expensive to decode.

@@

\subsection{Operators}
%todo: figure out how to do range queries from this paper somehow...
The operations supported are dictated by the Dictionary Matching problem. These operations are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{}, and \Range{}. The behavior of each operation for dynamic COSB-trees is described below. \Insertkonly{} and \Delete{} are also supported but not described here.

@@

\subsection{Complexity}
\Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high probability (w.h.p.). Traversing the centroid tree requires  $O(1+\log_{B}N)$ memory transfers, since the tree is built using$O(N/logN)$ keys and is balanced. Once in the hashdata layer, we are already within $O(logN)$ keys of $k$. Thus the local scan requires $O(1+(||k||+\log{}N)/B)$ memory transfers. Obtaining $k$ from the keydata layer requires no more memory transfers than the scan in the hashdata layer. The \Pred{} and \Succ{} operations are almost identical in behavior to searches, and thus only have the added cost of having to operate on $k'$, where $k'$ is the predecessor or successor, respectively. These operations as a result take $O(1+\log_{B}N+||k||/B+||k'||/B)$ memory transfers w.h.p.. Similarly, range queries take $O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory transfers, where in this case $k'$ is the other end of the range and $Q$ is the set of keys within the given range.

@@

\subsection{Motivation}
B-trees are often used as an indexing data structure for various
database and file systems. Thus, for a cache-oblivious B-tree data structure
to be of practical value, it must be able to handle concurrent access and modification
of its internal state. The straw-man solution is to simply ensure mutual exclusion
on the entire data structure; this approach, however, yields poor performance.

@@

\subsection{Description}
Before describing the data structures in more detail, we first consider
the concurrency model which is used throughout \cite{BenderFiGi05}.

\subsubsection{Concurrency Model}
The concurrency model starts with a system that has a total cache
size of $M$ and block size of $B$. There are $P$ processors in the
system, each with a cache of size $\frac{M}{P}$. A particular block
can reside in multiple caches in \textit{shared} mode. However, if a
processor wants to perform a write to a block, it must acquire it in 
\textit{exclusive} mode, in which case that block can only reside in
a single cache. In this model, a request for a block costs a single
memory transfer.

@@

\subsection{Exponential CO B-tree}
This data structure is based on a strongly weight-balanced 
exponential tree found in \cite{Bender2002}. A tuning parameter
$1 < \alpha < 2$ is chosen for the data structure. Each node
in the tree maintains both keys which divide up its children,
in addition to a \textit{right-link} pointer, which contains
both a pointer to a node's right sibling, plus the key
value of the minimum key in the node's right sibling. If no such
sibling exists, a sentinel $\infty$ pointer is used.

@@

\subsubsection{Operations}
The operations supported by the exponential CO B-tree are \Search{}
and \Insert{}. The authors do not discuss range-queries, and this
is presumably due to the concurrency issues in implementing a fully-consistent
range scan (weakly-consistent range scans could be implemented in the data
structure with very little modification), although interestingly enough
the PMA variant of B-tree (discussed in Section \ref{sec:pma}) does
cover range scans.

@@

\subsubsection{Complexity}
Sequential search takes $O(\log_B{N} + \log_{\alpha}{\log_2{B}})$ block
transfers (with high probability). The probabilistic bound comes from the
probabilistic key promotion property. The intuition behind the proof of this
bound comes from the argument that when the tree has height $h =
\Omega(\log_{\alpha}{\log_2{\log_2{N}}})$, then the total number of keys in the
tree is $O(2^{\alpha^h})$. The proof is then completed by summing over the
range of $h$ which is probabilistic $O(\log_{\alpha}{\log_2{N}})$, the access
time from some level $h$ to find the correct child.

@@

\subsection{Packed Memory Array CO B-tree}
\label{sec:pma}
The PMA CO B-tree is a two level data structure, containing a static
CO B-tree which indexes into a packed-memory array. Bender introduces
a new packed-memory array, called a \textit{one-way packed-memory array},
which they argue is easier to manage concurrently.

@@

% TODO: can somebody check this time-bound? seems kind of bad
Sequential search time-bounds are not presented in the paper,
but presumably since search is a static CO tree lookup followed
by a scan, the time bound is $O(\log_{B}{N} + {\log_2{N}}/{B} + 1)$.

@@

\subsubsection{Lock-free Variant}
The lock-free variant is based on the PMA CO tree, except the locks acquired
during the PMA modification are replaced by augmenting each element in the PMA
with a marker. A marker is used to denote the ongoing (concurrent) operation
of a particular cell.

>>>>

\caption{An illustration of the dynamic COSB-tree structure}
\label{fig:string_structure}
\end{figure}
\subsection{Motivation}

In order for B-trees to remain versatile, they must be able to accommodate
various forms of data. However, traditional B-trees and string B-trees have
suboptimal performance when variable length strings are used as keys. In
addition, string B-trees do not support compression, and commonly used front
compression is handled sub-optimally. Regular B-trees and string B-trees are
not cache oblivious, and thus incur poorer performance as described
previously. Range queries have poor performance in B-trees. In
\cite{BenderFaKu06}, Bender et al. address these issues by designing a Cache
Oblivious String B-tree (COSB-tree), which they argue works efficiently in
these problem areas for variable length string keys. They also provide a
modified front-compression scheme, which gives their approach the advantages
of using less space and having improved locality within the compressed data.

Two experiments are performed, in which a regular B-tree implementation tuned
to various block sizes is compared to a static COSB-tree implementation. The
experiment involved measuring the amount of time required to perform: 440,000
and 450,000 inserts of random values, a range query of all values, and 1000
random searches. The results of the experiment show that COSB-trees
outperformed B-trees most of the time for random searches, range queries and
random inserts. However, some point between 440,000 and 450,000 inserts the
COSB-tree had to rebalance the entire tree, causing it to fall behind in this
measurement. However, Bender et al. argue that the experiment was biased
towards the B-tree implementation, since the trees are "young", and thus still
well organized. The COSB-tree implementation is also compared to Berkeley DB
**cite berkeley db**, and achieves performance comparable to Berkeley DB with
tuned parameters. The results seem to show that there is a window in which
large and small-block size B-trees perform poorly, and that COSB-trees may be
a reasonable alternative to improve performance.

\subsubsection{General COSB-tree Structure}

The static COSB-tree consists of 2 layers: a centroid tree at the top, and an
array of the keys stored in lexicographic order on the bottom. The leaves of
the centroid tree contain pointers to keys in the array. The centroid tree is
structured such that a search through the centroid tree identifies the local
area of the desired key in the array. Then a sequential scan is performed
across the array to find the key.

@@

\caption{A centroid tree example}
\label{fig:centroid}
\end{figure}

In this approach to solving the dictionary matching problem *cite dictionary
matching problem paper*, a dictionary $D$ of $N$ keys is preprocessed using
hashing and divide-and-conquer via the construction of a centroid tree over
the keys. This technique forms the basis for the Cache Oblivious String B-tree
(COSB-tree) design. Consider the compacted trie $T$ of $D$. Bender et al.
observe that there exists a centroid vertex $\rho$ in $T$ such that $\rho$ has
between N/3 and 2N/3 descendants, inclusive. When searching for a key $k$ in
the compacted trie, you traverse the tree as follows: for each node $y$ that
you visit, if $y$ is a prefix of $x$, or $y$ matches $x$, continue on into the
trie rooted at $y$ or \emph{down trie} of $y$. Otherwise traverse the trie
resulting from excluding $y$ and its subtree. The \emph{centroid tree} is
obtained by rooting the tree at \emph{centroid vertex} $\rho$, and making its
children recursively defined centroid trees of its up and down tries.

@@

\subsubsection{Modifying Front Compression to Preserve Locality}

This section provides a brief summary of how front compression is modified to
optimize compression and decoding for static and dynamic cache-oblivious
string B-trees. Consider a sequence of $j$ keys to store $k_{1}$, $k_{2}$,
..., $k_{j}$. Without compression, this requires $\sum_{j}||k_{j}||$ space in
memory. To compress these keys using front compression, let $\pi_{j}$ be the
longest common prefix between $k_{j}$ and $k_{j+1}$. Now instead we store the
keys in the following format:

\begin{center}
$k_{1}$,$||\pi_{2}||$,$\sigma_{2}$,$||\pi_{3}||$$\sigma_{3}$,...,$||\pi_{j}||$,$\sigma_{j}$
\end{center}

where $\sigma_{j}$ is the suffix of $k_{j}$ after removing the first $\pi_{j}$
bits. This saves almost $\sum_{j}||\pi_{j}||$ memory. However, since decoding
$k_{j}$ involves concatenating the first $\pi_{j}$ bits of $k_{j-1}$ to
$\sigma_{j}$, $k_{j-1}$ must be decoded. $k_{j-1}$ may depend on previous keys
as well, making $k_{j}$ expensive to decode.

@@

\subsection{Operators}

%todo: figure out how to do range queries from this paper somehow...

The operations supported are dictated by the Dictionary Matching problem.
These operations are: \Search{} (synonymous with \Member{}), \Pred{}, \Succ{},
and \Range{}. The behavior of each operation for dynamic COSB-trees is
described below. \Insertkonly{} and \Delete{} are also supported but not
described here.

@@

\subsection{Complexity}

\Search{} takes $O(1+\log_{B}N+||k||/B)$ memory transfers with high
probability (w.h.p.). Traversing the centroid tree requires $O(1+\log_{B}N)$
memory transfers, since the tree is built using$O(N/logN)$ keys and is
balanced. Once in the hashdata layer, we are already within $O(logN)$ keys of
$k$. Thus the local scan requires $O(1+(||k||+\log{}N)/B)$ memory transfers.
Obtaining $k$ from the keydata layer requires no more memory transfers than
the scan in the hashdata layer. The \Pred{} and \Succ{} operations are almost
identical in behavior to searches, and thus only have the added cost of having
to operate on $k'$, where $k'$ is the predecessor or successor, respectively.
These operations as a result take $O(1+\log_{B}N+||k||/B+||k'||/B)$ memory
transfers w.h.p.. Similarly, range queries take
$O(1+\log_{B}N+(||k||+||k'||+\llangle{}Q\rrangle{})/B)$ memory transfers,
where in this case $k'$ is the other end of the range and $Q$ is the set of
keys within the given range.

@@

\subsection{Motivation}

B-trees are often used as an indexing data structure for various database and
file systems. Thus, for a cache-oblivious B-tree data structure to be of
practical value, it must be able to handle concurrent access and modification
of its internal state. The straw-man solution is to simply ensure mutual
exclusion on the entire data structure; this approach, however, yields poor
performance.

@@

Before describing the data structures in more detail, we first consider
the concurrency model which is used throughout \cite{BenderFiGi05}.

\subsubsection{Concurrency Model}

The concurrency model starts with a system that has a total cache size of $M$
and block size of $B$. There are $P$ processors in the system, each with a
cache of size $\frac{M}{P}$. A particular block can reside in multiple caches
in \textit{shared} mode. However, if a processor wants to perform a write to a
block, it must acquire it in \textit{exclusive} mode, in which case that block
can only reside in a single cache. In this model, a request for a block costs
a single memory transfer.

@@

\subsection{Exponential CO B-tree}

This data structure is based on a strongly weight-balanced exponential tree
found in \cite{Bender2002}. A tuning parameter $1 < \alpha < 2$ is chosen for
the data structure. Each node in the tree maintains both keys which divide up
its children, in addition to a \textit{right-link} pointer, which contains
both a pointer to a node's right sibling, plus the key value of the minimum
key in the node's right sibling. If no such sibling exists, a sentinel
$\infty$ pointer is used.

@@

The operations supported by the exponential CO B-tree are \Search{} and
\Insert{}. The authors do not discuss range-queries, and this is presumably
due to the concurrency issues in implementing a fully-consistent range scan
(weakly-consistent range scans could be implemented in the data structure with
very little modification), although interestingly enough the PMA variant of
B-tree (discussed in Section \ref{sec:pma}) does cover range scans.

@@

Sequential search takes $O(\log_B{N} + \log_{\alpha}{\log_2{B}})$ block
transfers (with high probability). The probabilistic bound comes from the
probabilistic key promotion property. The intuition behind the proof of this
bound comes from the argument that when the tree has height $h =
\Omega(\log_{\alpha}{\log_2{\log_2{N}}})$, then the total number of keys in
the tree is $O(2^{\alpha^h})$. The proof is then completed by summing over the
range of $h$ which is probabilistic $O(\log_{\alpha}{\log_2{N}})$, the access
time from some level $h$ to find the correct child.

@@

\subsection{Packed Memory Array CO B-tree}
\label{sec:pma}

The PMA CO B-tree is a two level data structure, containing a static CO B-tree
which indexes into a packed-memory array. Bender introduces a new
packed-memory array, called a \textit{one-way packed-memory array}, which they
argue is easier to manage concurrently.

@@

% TODO: can somebody check this time-bound? seems kind of bad

Sequential search time-bounds are not presented in the paper, but presumably
since search is a static CO tree lookup followed by a scan, the time bound is
$O(\log_{B}{N} + {\log_2{N}}/{B} + 1)$.

@@

\subsubsection{Lock-free Variant}

The lock-free variant is based on the PMA CO tree, except the locks acquired
during the PMA modification are replaced by augmenting each element in the PMA
with a marker. A marker is used to denote the ongoing (concurrent) operation
of a particular cell.
